---
title: "Pre-processing of sample data"
author: "Andrea Villanueva Raisman"
output: html_document
date: "2023-12-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
#Import libraries
library(tidyverse)
library(readxl)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(modelr)
library(devtools)
library(HarmonizR)
library(dplyr)
```


## Importing filtered plate reports

In this section, the data file for all plates will be imported after having filtered out the peptides with a dotp of under 80 and the samples with under 75 good quality proteins and log10-transformed ratios to the standard of under -0.5. This filtered data, which contains the mass spectrometry results, will be merged with the file that contains the identification code for each sample. 

```{r, echo=FALSE}

filtered_data <- read.csv("filtered_data.csv")

filtered_data$Position <- filtered_data$`plate_well`
filtered_data <- filtered_data[, !grepl("^plate_well", names(filtered_data))]
filtered_data <- filtered_data %>%
  filter(IsHeavy == FALSE)
# Read file containing identification code for each sample
sample_identification <- read_excel("220411_SampleManifest_Olink_DA3.xlsx") 
sample_identification <- sample_identification[c("target_plate", "target_pos", "sample_id")]

# Combine 'plate' and 'well' columns into a new column 'plate_well' 
sample_identification$Position <- paste(sample_identification$target_plate, sample_identification$target_pos, sep = "_")

# Drop the Plate and Well columns in sample_identification
sample_identification <- sample_identification[, !grepl("^target_plate|^target_pos", names(sample_identification))]

# Merge data frames based on the new combined column 'plate_well'
Identified_data <- merge(filtered_data, sample_identification, by = "Position")

Identified_data <- Identified_data[, c("Plate", "Well", "Position", "sample_id", "Protein", "Peptide", "Ratio.To.Standard", "DotProductLightToHeavy", "Total.Area" )]

```

# Distribution and outliers

```{r, Histogram, echo=FALSE}
#Histogram of normalized area
ggplot(Identified_data, aes(x = log10(Total.Area))) +
  geom_histogram(fill = 'lightpink', color = 'pink2', bins = 100) +
  labs(title = "Total area distribution",
       x = "Log10 of total area",
       y = "Frequency") +
  theme_bw()
```
```{r, boxplots, echo=FALSE}

plate_colors <- c("3" = "darkred", "4" = "red2", "6"="red", "7"="darkorange", "9"="orange", "10"="brown","15" ="gold3", "16"="gold", "18"="green3", "21"="green4", "22" = "turquoise", "24" = "darkturquoise", "25"= "lightblue1", "27"="lightblue", "29"="blue", "30"="blue3", "31"="darkblue", "32"="purple", "33"="purple4" )

# Convert Plate to a factor for proper ordering
Identified_data$Plate <- factor(Identified_data$Plate)



# Boxplot with facets by Plate and x-axis labels as sample_id
suppressWarnings(
  ggplot(Identified_data, aes(x = Plate, y = `Ratio.To.Standard`)) + 
  geom_boxplot(aes(fill = Plate), alpha = 0.6, outlier.shape = NA, width = 0.7) +  # Adjust width for boxplot
  geom_jitter(aes(color = Plate), width = 0.1, height = 0.1, alpha = 0.3, size = 0.5, stroke = 0.5) +  # Adjust jitter aesthetics
  scale_fill_manual(values = plate_colors) +
  scale_color_manual(values = plate_colors) +  # Set colors for jittered points
  stat_summary(fun = "mean",
               geom = "crossbar",
               color = "black",
               width = 0.5, show.legend = FALSE) +
      theme(
        title = element_text(size = 12),
        axis.title.y = element_text(size = 10, margin = margin(t = 0, r = 10, b = 0, l = 0)), 
        axis.title.x = element_text(size = 10, margin = margin(t = 10, r = 0, b = 0, l = 0)), 
        panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "lightgrey"),
        axis.text.x = element_text(angle = 90),
        legend.position = "none"
      ) +
    ylim(0,2000) +
      labs(title = "Ratio to standard", x = "Plates", y ="Ratio to standard")  # Add title and axis labels
)
```

```{r, echo=FALSE}
#APOA1
apoa1_data <- subset(Identified_data, Protein == "sp|P02647|APOA1_HUMAN")


#Set peptide colors 
peptide_colors <- c("VSFLSALEEYTK" = "#FF0000", "QGLLPVLESFK" = "#FF6666")

#This together with the last line of the plot ensures that all plate numbers are visible
apoa1_data$Plate <- factor(apoa1_data$Plate)

#Visualize ratio to standard
ggplot(apoa1_data, aes(x = sample_id, y = `Ratio.To.Standard`)) +
  geom_point(aes(color = `Peptide`)) +
  labs(title = "APOA1 ratio to standard", x = "Plate", y = "Ratio to standard") +
  scale_color_manual(values = peptide_colors) +
  facet_wrap(~ Plate, nrow = 1) +
  theme_bw()  +
  scale_x_discrete(labels = levels(apoa1_data$Plate))+
  theme(axis.text.x = element_blank()) +
  ylim(min(apoa1_data$Ratio.To.Standard), max(apoa1_data$Ratio.To.Standard))
```

## Batch effect correction

After the batch effect analysis using the in-house plasma samples in each plate. We perform median normalization and batch effect correction for all samples. The first step is performing median centering and we can compare the effect in a boxplot before and after.

```{r, echo=FALSE}
# Median centering within each plate for each peptide
median_centered_data <- Identified_data %>%
  group_by(Plate, Peptide) %>%
  mutate(`Ratio.To.Standard` = `Ratio.To.Standard` - median(`Ratio.To.Standard`))
```


```{r, echo=FALSE}
# Convert Plate to a factor for proper ordering
median_centered_data$Plate <- factor(median_centered_data$Plate)

# Convert Plate to a factor for proper ordering
median_centered_data$Plate <- factor(median_centered_data$Plate)

# Boxplot with facets by Plate and x-axis labels as sample_id
suppressWarnings(
  ggplot(median_centered_data, aes(x = Plate, y = `Ratio.To.Standard`)) + 
  geom_boxplot(aes(fill = Plate), alpha = 0.6, outlier.shape = NA, width = 0.7) +  # Adjust width for boxplot
  geom_jitter(aes(color = Plate), width = 0.1, height = 0.1, alpha = 0.3, size = 0.5, stroke = 0.5) +  # Adjust jitter aesthetics
  scale_fill_manual(values = plate_colors) +
  scale_color_manual(values = plate_colors) +  # Set colors for jittered points
  stat_summary(fun = "mean",
               geom = "crossbar",
               color = "black",
               width = 0.5, show.legend = FALSE) +
      theme(
        title = element_text(size = 12),
        axis.title.y = element_text(size = 10, margin = margin(t = 0, r = 10, b = 0, l = 0)), 
        axis.title.x = element_text(size = 10, margin = margin(t = 10, r = 0, b = 0, l = 0)), 
        panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "lightgrey"),
        axis.text.x = element_text(angle = 90),
        legend.position = "none"
      ) +
    ylim(-1000,1000) +
      labs(title = "Ratio to standard", x = "Plates", y ="Ratio to standard")  # Add title and axis labels
)
```

```{r, echo=FALSE}
#APOA1
apoa1_data <- subset(median_centered_data, Protein == "sp|P02647|APOA1_HUMAN")

#Set peptide colors 
peptide_colors <- c("VSFLSALEEYTK" = "#FF0000", "QGLLPVLESFK" = "#FF6666")

#This together with the last line of the plot ensures that all plate numbers are visible
apoa1_data$Plate <- factor(apoa1_data$Plate)

#Visualize ratio to standard
ggplot(apoa1_data, aes(x = sample_id, y = `Ratio.To.Standard`)) +
  geom_point(aes(color = `Peptide`)) +
  labs(title = "Median centered APOA1 ratio to standard", x = "Plate", y = "Ratio to standard") +
  scale_color_manual(values = peptide_colors) +
  facet_wrap(~ Plate, nrow = 1) +
  theme_bw()  +
  scale_x_discrete(labels = levels(apoa1_data$Plate))+
  theme(axis.text.x = element_blank()) +
  ylim(min(apoa1_data$Ratio.To.Standard), max(apoa1_data$Ratio.To.Standard))
```

```{r}
# Saving the filtered dataframe to a CSV file
write.csv(median_centered_data, file = "median_centered_data.csv", row.names = FALSE)
```