---
title: "Quality control for plates"
author: "Andrea Villanueva Raisman"
output: html_document
---
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE}
library(tidyverse)
library(readxl)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(modelr)
library(gridExtra)
library(dplyr)
library(kableExtra)
```


```{r, echo=FALSE}

# Function to read and process plate reports
read_process_plate <- function(plate_data) {

  plate_data <- read.csv(plate_data)
  
  #Get rid of QTags
  plate_data <- subset(plate_data, Protein != "QTag")
  
  #Transform from character to numeric class
  plate_data$Ratio.To.Standard <- as.numeric(sub(",", ".", plate_data$Ratio.To.Standard), na.rm = TRUE)
  plate_data$DotProductLightToHeavy <- as.numeric(sub(",", ".", plate_data$DotProductLightToHeavy), na.rm = TRUE)
  #plate_data$Total.Area <- as.numeric(sub(",", ".", plate_data$Total.Area))
  
  #Separation of plate (batch) and wells
  plate_data <- separate(plate_data, Replicate.Name, into = c("Plate", "Well"), sep = "_")
  plate_data$Plate <- as.numeric(sub(",", ".", plate_data$Plate))
  
  # Adjust well names (remove first letter)
  plate_data$Well <- substr(plate_data$Well, 2, 4)  # Keep only last three digits
  
  return(plate_data)
}

# File paths for plate reports 
plate_files <- c("plate_3 report.csv", "plate_4 report.csv", "plate_6 report.csv", "plate_7 report.csv", "plate_9 report.csv", "plate_10 report.csv", "plate_15 report.csv", "plate_16 report.csv", "plate_21 report.csv", "plate_22 report.csv", "plate_24 report.csv", "plate_25 report.csv", "plate_27 report.csv", "plate_29 report.csv", "plate_30 report.csv", "plate_31 report.csv", "plate_32 report.csv", "plate_33 report.csv")

# Import and process each plate report
plate_reports <- lapply(plate_files, read_process_plate)

combined_data <-  bind_rows(plate_reports)
```

# Data distribution and outliers

In the first histogram, by looking at the normalized area of the peptide intensity peaks, we can assess if they follow a normal distribution. All histograms shows a skewed central tendency because of the many, very low values, which also indicate potential outliers. However, there do not seem to be gaps in the data for any of the plates. The boxplots facilitate the identification of the outliers and display their relative distance from the bulk of the data, making it easier to compare distribution across the samples and the plates. 

```{r, raw_histograms, height = 24, width = 12, echo=FALSE}

plate_colors <- c("3" = "darkred", "4" = "red2", "6"="red", "7"="darkorange", "9"="orange", "10"="brown","15" ="gold3", "16"="gold", "18"="green3", "21"="green4", "22" = "turquoise", "24" = "darkturquoise", "25"= "lightblue1", "27"="lightblue", "29"="blue", "30"="blue3", "31"="darkblue", "32"="purple", "33"="purple4" )

combined_data$Plate <- factor(combined_data$Plate)

plot_data <- data.frame(combined_data)
  
suppressWarnings(ggplot(plot_data, aes(x = log10(`Total.Area`))) +
      geom_histogram(aes(fill = Plate), bins = 100, alpha = 0.8) +
      facet_wrap(~ Plate, nrow = 2, scales = "fixed") +
      scale_fill_manual(values = plate_colors) +
      theme(
        title = element_text(size = 12),
        axis.title.y = element_text(size = 10, margin = margin(t = 0, r = 10, b = 0, l = 0)), 
        axis.title.x = element_text(size = 10, margin = margin(t = 10, r = 0, b = 0, l = 0)), 
        panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "lightgrey"),
        axis.text.x = element_text(angle = 90),
        legend.position = "none"
      ) +
      labs(title = "Total peptide area distribution", x = "log10 of total peptide area", y = "Frequency")  # Add title and axis labels

)
```

The outliers shown by the boxplot can be the result of bad peak integration. In some cases, this can be the result of incorrect peak integration which can be corrected manually. Skyline's results have not been manually revised for this plate yet due to time constraints, but the most extreme outliers most probably belong to heavy isotopes which have not detected. 

```{r, raw_boxplots, height = 24, width = 12, echo=FALSE}

suppressWarnings(
ggplot(plot_data, aes(x = Well, y = log10(`Ratio.To.Standard`))) +
  geom_boxplot(aes(fill = Plate), alpha = 0.6, outlier.shape = NA, width = 0.8) +  # Adjust width for boxplot
  geom_jitter(aes(color = Plate), width = 0.1, height = 0.1, alpha = 0.3, size = 0.5, stroke = 0.5) +  # Adjust jitter aesthetics
  facet_wrap(~ Plate, nrow = 2, scales = "fixed") +
  scale_fill_manual(values = plate_colors) +
  scale_color_manual(values = plate_colors) +  # Set colors for jittered points
  stat_summary(fun = "mean",
               geom = "crossbar",
               color = "black",
               width = 0.5, show.legend = FALSE) +
  theme(
        title = element_text(size = 12),
        axis.title.y = element_text(size = 10, margin = margin(t = 0, r = 10, b = 0, l = 0)), 
        axis.title.x = element_text(size = 10, margin = margin(t = 10, r = 0, b = 0, l = 0)), 
        panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "lightgrey"),
        axis.text.x = element_blank(),
        legend.position = "none"
      ) +
      labs(title = "Ratio to standard", x = "Wells", y ="Log10 of ratio to standard")  # Add title and axis labels
)
```

We proceed by filtering the peptides whose dotp product is low. The thresold here is dotp = 0.80. However, the report might be improved later after manually re-integrating peaks. The histograms now show a less skewed central tendency because many of the low values were due to poor integration. If this is not aparent in the histograms, the table shows the numerical values of the median of the total area per plate and the difference between them. The third column is the result of substracting the median of the raw data from the filtered data. In all cases, the difference is positive, meaning that filtering the data increased the median in all plates.

```{r, echo=FALSE}
#Filter peptides with low dotp
filtered_data <- combined_data %>%
  filter(`DotProductLightToHeavy` > 0.80)
```

```{r, filtered_histograms, height = 24, width = 12, echo=FALSE}

filtered_data$Plate <- factor(filtered_data$Plate)

plot_data_f <- data.frame(filtered_data)
  
suppressWarnings(ggplot(plot_data_f, aes(x = log10(`Total.Area`))) +
      geom_histogram(aes(fill = Plate), bins = 100, alpha = 0.8) +
      facet_wrap(~ Plate, nrow = 2, scales = "fixed") +
      scale_fill_manual(values = plate_colors) +
      theme(
        title = element_text(size = 12),
        axis.title.y = element_text(size = 10, margin = margin(t = 0, r = 10, b = 0, l = 0)), 
        axis.title.x = element_text(size = 10, margin = margin(t = 10, r = 0, b = 0, l = 0)), 
        panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "lightgrey"),
        axis.text.x = element_text(angle = 90),
        legend.position = "none"
      ) +
      labs(title = "Total peptide area distribution", x = "log10 of total peptide area", y = "Frequency")  # Add title and axis labels

)

```

**Median comparison table: before and after low dotp filtering**
```{r, echo=FALSE}
# Calculate median Total Area per Plate for combined_data
combined_medians <- combined_data %>%
  group_by(Plate) %>%
  summarize(median_Total_Area = median(`Total.Area`, na.rm = TRUE))

# Calculate median Total Area per Plate for filtered_data
filtered_medians <- filtered_data %>%
  group_by(Plate) %>%
  summarize(median_Total_Area = median(`Total.Area`, na.rm = TRUE))

# Merge the medians for comparison
comparison_table <- merge(combined_medians, filtered_medians, by = "Plate", suffixes = c("_before", "_after"))


# Calculate the difference between medians (optional)
comparison_table$Median_difference <- comparison_table$median_Total_Area_after - comparison_table$median_Total_Area_before

colnames(comparison_table) <- c("Plate", "Raw data median", "Filtered data median", "Median difference")

# Print the comparison table
print(comparison_table)

```

Similarly, the most extreme outliers have disappeared from the boxplots. What now becomes visible is that there are samples with much lower medians than the rest in all plates too. However, the number of low-median samples per plate dramatically increses in plates 30 to 33. This might be due to the failure detected in the robotics equipment around the time that these samples were being prepared for the mass spectrometer.

```{r, height = 24, width = 12, echo=FALSE}
suppressWarnings(
ggplot(plot_data_f, aes(x = Well, y = log10(`Ratio.To.Standard`))) +
  geom_boxplot(aes(fill = Plate), alpha = 0.6, outlier.shape = NA, width = 0.8) +  # Adjust width for boxplot
  geom_jitter(aes(color = Plate), width = 0.1, height = 0.1, alpha = 0.3, size = 0.2, stroke = 0.5) +  # Adjust jitter aesthetics
  facet_wrap(~ Plate, nrow = 2, scales = "fixed") +
  scale_fill_manual(values = plate_colors) +
  scale_color_manual(values = plate_colors) +  # Set colors for jittered points
  stat_summary(fun = "mean",
               geom = "crossbar",
               color = "black",
               width = 0.5, show.legend = FALSE) +
 theme(
        title = element_text(size = 12),
        axis.title.y = element_text(size = 10, margin = margin(t = 0, r = 10, b = 0, l = 0)), 
        axis.title.x = element_text(size = 10, margin = margin(t = 10, r = 0, b = 0, l = 0)), 
        panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "lightgrey"),
        axis.text.x = element_blank(),
        legend.position = "none"
      ) +
      labs(title = "Ratio to standard",  x = "Wells", y ="Log10 of ratio to standard")
)
```


## Counts

Looking at the count of heavy standards versus endogenous peptides also allows to make an assessment of where potential errors during sample preparation might lie, especially if there is a difference between the detected heavy and light isotopes in a sample. At the same time, samples with low peptide counts can be discarded, at least temporarily because they indicate that a process that introduced noise might have taken place. The following plots show the correlation between light and heavy isotopes. It should be noted that plates 4 and 21 had samples with peptide counts of over 400. Since there are only 257 peptides in the target list, this must be manually revised. All the plates show a nearly perfect positive correlation between detected light and heavy isotopes.   

```{r, echo=FALSE}
# Create Light_Isotopes column based on conditions
filtered_data$Count <- ifelse(!is.na(filtered_data$Total.Area) & filtered_data$Total.Area != 0, 1, 0)

# Creating a new column 'IsHeavy' indicating if the peptide is heavy
filtered_data$IsHeavy <- grepl("\\(heavy\\)$", filtered_data$Precursor)

# Grouping by 'Replicate' and 'IsHeavy' to count detected peptides
count_summary <- aggregate(Count ~ interaction(Plate, Well) + IsHeavy, data = filtered_data, FUN = sum)

# Reshaping data to wide format
count_summary_wide <- reshape(count_summary, idvar = "interaction(Plate, Well)", timevar = "IsHeavy", direction = "wide")

count_summary_wide <- separate(count_summary_wide, `interaction(Plate, Well)`, into = c("Plate", "Well"))

count_summary_wide$Plate <- as.factor(as.numeric(count_summary_wide$Plate))

# Renaming columns
colnames(count_summary_wide) <- c("Plate", "Well", "LightPeptides", "HeavyPeptides")

```

```{r, height = 100, width = 25, echo=FALSE}
# Create a scatter plot
suppressWarnings(
ggplot(count_summary_wide, aes(x = LightPeptides, y = HeavyPeptides)) +
  geom_point(aes(fill = Plate), alpha = 0.6, width = 0.8) +  # Adjust width for boxplot
  geom_jitter(aes(color = Plate), width = 0.1, height = 0.1, alpha = 0.3, size = 0.5, stroke = 0.5) +  # Adjust jitter aesthetics
  facet_wrap(~ Plate, nrow = 6) +
  scale_fill_manual(values = plate_colors) +
  scale_color_manual(values = plate_colors) +  # Set colors for jittered points
 theme(
        title = element_text(size = 12),
        axis.title.y = element_text(size = 10, margin = margin(t = 0, r = 10, b = 0, l = 0)), 
        axis.title.x = element_text(size = 10, margin = margin(t = 10, r = 0, b = 0, l = 0)), 
        panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "lightgrey"),
        axis.text.x = element_text(size=6),
        axis.text.y = element_text(size=6),
        legend.position = "none"
      ) +
      ylim(0,257) +
      xlim(0,257) +
      labs(title = "Light vs heavy isotope count",  x = "Light peptides", y ="Heavy peptides")
)
```

The subsequent step in the analysis involves filtering samples based on two criteria: peptide count and the median of their ratios to the standard. This dual filtration approach is adopted to ensure a robust selection of samples that exhibit both a significant number of detected peptides and a consistent ratio to the standard across proteins.

The criteria involve retaining samples from the plate reports where a minimum of 75 out of 117 proteins were detected. Additionally, we filter samples based on a criterion where the median of the log10-transformed ratios to the standard exceeds -0.5. Although the protein count threshold predominantly filters out samples with lower medians, some samples with medians close to this threshold may persist. Filtering by median values allows to specifically target and exclude these samples, ensuring a more stringent selection.

Employing both filters serves as a means of reinforcing the selection process. The combined approach ensures that samples exhibiting a low peptide count or potentially deviating median values are excluded, which could otherwise introduce bias or skewness in subsequent analyses. Moreover, employing only the peptide count as a filter might retain samples with a higher number of detected peptides but significantly lower peptide intensities. Such samples may reflect non-biological processes or technical artifacts, warranting further investigation in subsequent analyses.

Visualizing the resulting filtered dataset using the same boxplot reveals a more regular distribution among the retained samples. This filtration process helps in ensuring a dataset with a more consistent and reliable distribution for subsequent analyses.

```{r, echo=FALSE}
# Calculate the count of unique proteins per well
filtered_data$plate_well <- paste(filtered_data$Plate, filtered_data$Well, sep = "_")

protein_count <- filtered_data %>%
  group_by(plate_well) %>%
  summarise(`Unique Proteins` = n_distinct(Protein))

# Set the threshold count of proteins
threshold <- 75

# Filter wells where the UniqueProteins count is at least 66% of total proteins
selected_wells <- protein_count %>%
  filter(`Unique Proteins` >= threshold) %>%
  pull(plate_well)

# Filter the original dataframe based on the selected wells
QC_data <- filtered_data%>%
  filter((plate_well) %in% selected_wells) %>%
  group_by(plate_well) %>%
  filter(log10(median(Ratio.To.Standard)) >= -0.5)
```

```{r, height = 24, width = 12, echo=FALSE}

QC_data$Plate <- factor(QC_data$Plate)

plot_data_QC <- data.frame(QC_data)
  
suppressWarnings(
ggplot(plot_data_QC, aes(x = Well, y = log10(`Ratio.To.Standard`))) +
  geom_boxplot(aes(fill = Plate), alpha = 0.6, outlier.shape = NA, width = 0.8) +  # Adjust width for boxplot
  geom_jitter(aes(color = Plate), width = 0.1, height = 0.1, alpha = 0.3, size = 0.2, stroke = 0.5) +  # Adjust jitter aesthetics
  facet_wrap(~ Plate, nrow = 2, scales = "fixed") +
  scale_fill_manual(values = plate_colors) +
  scale_color_manual(values = plate_colors) +  # Set colors for jittered points
  stat_summary(fun = "mean",
               geom = "crossbar",
               color = "black",
               width = 0.5, show.legend = FALSE) +
 theme(
        title = element_text(size = 12),
        axis.title.y = element_text(size = 10, margin = margin(t = 0, r = 10, b = 0, l = 0)), 
        axis.title.x = element_text(size = 10, margin = margin(t = 10, r = 0, b = 0, l = 0)), 
        panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "lightgrey"),
        axis.text.x = element_blank(),
        legend.position = "none"
      ) +
      labs(title = "Ratio to standard",  x = "Wells", y ="Log10 of ratio to standard")
)
```


The conclusions that can be extracted from this preliminary exploration are:
    1. The original data had a more skewed central tendency towards lower values for all plates, with some very extreme outliers some. Those outliers which were the result of low quality peaks were removed by filtering the data using the threshold of dotp = 0.80. 
    2. After filtering out bad quality measurements, the median of the total area per plate increased in all cases and the most extreme outliers disappeared. 
    3. All plates had samples whose median was much lower than the rest. However, plates 30, 31, 32 and 33 had a much higher number of these low-median samples than the rest, as can be seen in the boxplot for the filtered data.
    4. The correlation between light and heavy isotope detection is perfectly positive for all plates. The group in the bottom left corner in most cases was made up of the very same samples that had shown low medians in the boxplots. This is not shown here but was done by analyzing each plate individually.
    5. Most low-median samples were filtered out by using a threshold of at least 75 proteins out of 117. However, a second filter was used for those low-median samples that remained. The combination of both criteria should ensure a more robust selection of good-quality samples to continue for further analyses.


```{r, echo=FALSE}
# Saving the filtered dataframe to a CSV file
write.csv(QC_data, file = "filtered_data.csv", row.names = FALSE)
```

